\(ler\) : LVK (LIGO-Virgo-KAGRA collaboration) Event (compact-binary mergers) Rate calculator and simulator

### 2024-06-20 paper/paper.tex
mac

\(ler\) is a statistical-based Python package specifically designed for
computing detectable rates of both lensed and unlensed gravitational
wave (GW) events, catering to the requirements of the LIGO-Virgo-KAGRA
Scientific Collaboration and astrophysics research scholars. The core
functionality of \(ler\) intricately hinges upon the interplay of
various components which include sampling the properties of
compact-binary sources, lens galaxies characteristics, solving lens
equations to derive properties of resultant images, and computing
detectable GW rates. This comprehensive functionality builds on the
leveraging of array operations and linear algebra from the \emph{numpy}
library, enhanced by interpolation methods from \emph{scipy} and
Python's \emph{multiprocessing} capabilities. Efficiency is further
boosted by the \emph{numba} library's Just-In-Time (\emph{njit})
compilation, optimizing extensive numerical computations and employing
the inverse transform sampling method to replace more cumbersome
rejection sampling. The modular design of \(ler\) not only optimizes
speed and functionality but also ensures adaptability and upgradability,
supporting the integration of additional statistics as research evolves.
Currently, \(ler\) is an important tool in generating simulated
gravitational wave (GW) events---both lensed and unlensed---and provides
astrophysically accurate distributions of event-related parameters for
both detectable and non-detectable events. This functionality aids in
event validation and enhances the forecasting of detection capabilities
across various GW detectors to study such events. The architecture of
the \(ler\) API facilitates seamless compatibility with other software
packages, allowing researchers to integrate and utilize its
functionalities based on specific scientific requirements.

@@

  \begin{itemize}
  \tightlist
  \item
    For the unlensed events, The sampling distribution
    \((\,R_m(z_s)\,)\) for the source's redshift \((\,z_s\,)\) is
    derived from the merger rate density of compact binaries, which, in
    turn, is based on the star formation rate. The code is meticulously
    designed to enable the straightforward integration of future updates
    or user-specified distributions of these sources.
  \item
    The process of sampling both intrinsic and extrinsic parameters of
    GW sources, represented by \(\theta\), utilizes the prior
    distributions \((P(\theta))\) available within the \emph{gwcosmo}
    and \emph{bilby} Python packages. Additionally, users have the
    option to initialize the package with custom distributions of their
    choosing.
  \end{itemize}
\item
  Sampling of lens galaxies attributes and source red-shifts:

  \begin{itemize}
  \tightlist
  \item
    For lensed case, the source redshift \((z_s)\) is sampled under the
    strong lensing condition \((\text{SL})\), based on the precomputed
    probability of strong lensing with source at \(z_s\)
    \((\text{optical depth: }P\left(\text{SL}|z_s\right)\,)\). This
    probability can be recalculated for specified configurations of lens
    galaxies, leveraging \emph{multiprocessing} and \emph{njit}
    functionalities for enhanced efficiency.
  \item
    The package utilizes the Elliptical Power Law with external shear
    (EPL+Shear) model (Wempe et al. 2022) for galaxy parameters
    \((\theta_L)\) sampling, following reference
    (https://arxiv.org/abs/2106.06303)(Wierda et al. 2021). Rejection
    sampling is applied on the above samples on condition that whether
    the event is strongly lensed or not,
    \(P\left(\text{SL}|z_s,\theta_L\right)\).
  \end{itemize}
\item
  Generation of image properties:

  \begin{itemize}
  \tightlist
  \item
    Source position \((\beta)\) is sampled from the caustic in the
    source plane.
  \item
    Sampled lens properties and source position is fed in
    \emph{lenstronomy} to generate properties of the images. This is the
    slowest part of the entire simulation, which \(ler\) tackles through
    parallelization with multiprocessing.
  \item
    Image properties like magnification \((\mu_i)\) and time delay
    (\(\Delta t_i\)) modifies the original source signal strength,
    changing the signal-to-noise ratio SNR and our ability to detect.
  \end{itemize}
\item
  Calculation of Detectable Merger Rates Per Year:

  \begin{itemize}
  \tightlist
  \item
    The calculation of rates necessitates integration over simulated
    events that meet specific detection criteria. This process includes
    computing SNRs \((\rho)\) for each event or its lensed images,
    followed by an assessment against a predetermined threshold(s)
    \((\rho_{th})\).
  \item
    SNR calculations are optimized using
    \href{https://github.com/hemantaph/gwsnr}{\emph{gwsnr}}, leveraging
    interpolation and multiprocessing for accuracy and speed.
  \item
    Simulated events and rate results, along with input configurations,
    are systematically archived for easy access and future analysis.
    Additionally, all interpolators used in the process are preserved
    for future applications.
  \end{itemize}
\end{itemize}

@@

\begin{equation*}
\begin{split}
R_U = \int & dz_s \frac{dV_c}{dz_s}\frac{R_m(z_s)}{1+z_s}\left\{\Theta[\rho(z_s,\theta)-\rho_{th}] P(\theta) d\theta \right\}
\end{split}
\end{equation*}

@@

\begin{equation*}
\begin{split}
R_L = \int & dz_s \frac{dV_c}{dz_s}\tau(z_s)\frac{R_m(z_s)}{1+z_s} \,\mathcal{O}_{images}(z_s,\theta,\mu_i,\Delta t_i, \rho_{th}) \, \\ 
& \, P(\theta) P(\theta_L|\text{SL},z_s) P(\beta|\text{SL}) d\theta d\beta d\theta_L dz_s 
\end{split}
\end{equation*}

@@

\phantomsection\label{refs}
\begin{CSLReferences}{1}{0}
\bibitem[\citeproctext]{ref-Janquart2023}
Janquart, J, M Wright, S Goyal, J C L Chan, A Ganguly, Á Garrón, D
Keitel, et al. 2023. {``Follow-up Analyses to the O3 LIGO--Virgo--KAGRA
Lensing Searches.''} \emph{Monthly Notices of the Royal Astronomical
Society} 526 (3): 3832--60.
\url{https://doi.org/10.1093/mnras/stad2909}.

>>>>

\(ler\) is a statistical-based Python package specifically designed for
computing detectable rates of both lensed and unlensed gravitational
wave (GW) events, catering to the requirements of the LIGO-Virgo-KAGRA
Scientific Collaboration and astrophysics research scholars. The core
functionality of \(ler\) intricately hinges upon the interplay of
various components which include sampling the properties of
compact-binary sources, lens galaxies characteristics, solving lens
equations to derive properties of resultant images, and computing
detectable GW rates. This comprehensive functionality builds on the
leveraging of array operations and linear algebra from the \emph{numpy}
(NumPy Community 2022) library, enhanced by interpolation methods from
\emph{scipy} (Virtanen et al. 2020) and Python's \emph{multiprocessing}
capabilities. Efficiency is further boosted by the \emph{numba} (Lam,
Pitrou, and Seibert 2022) library's Just-In-Time (\emph{njit})
compilation, optimizing extensive numerical computations and employing
the inverse transform sampling method to replace more cumbersome
rejection sampling. The modular design of \(ler\) not only optimizes
speed and functionality but also ensures adaptability and upgradability,
supporting the integration of additional statistics as research evolves.
Currently, \(ler\) is an important tool in generating simulated
gravitational wave (GW) events---both lensed and unlensed---and provides
astrophysically accurate distributions of event-related parameters for
both detectable and non-detectable events. This functionality aids in
event validation and enhances the forecasting of detection capabilities
across various GW detectors to study such events. The architecture of
the \(ler\) API facilitates seamless compatibility with other software
packages, allowing researchers to integrate and utilize its
functionalities based on specific scientific requirements.

@@

  \begin{itemize}
  \tightlist
  \item
    For the unlensed events, The sampling distribution
    \((\,R_m(z_s)\,)\) for the source's redshift \((\,z_s\,)\) is
    derived from the merger rate density of compact binaries, which, in
    turn, is based on the star formation rate. The code is meticulously
    designed to enable the straightforward integration of future updates
    or user-specified distributions of these sources.
  \item
    Intrinsic and extrinsic parameters \((\theta)\) of GW sources are
    sampled using prior distributions \((P(\theta))\) from the gwcosmo
    (gwcosmo Contributors 2022) and bilby (\textbf{Ashton2019?})
    packages, with options for users to input custom distributions.
  \end{itemize}
\item
  Sampling of lens galaxies attributes and source red-shifts:

  \begin{itemize}
  \tightlist
  \item
    For lensed case, the source redshift \((z_s)\) is sampled under the
    strong lensing condition \((\text{SL})\), based on the precomputed
    probability of strong lensing with source at \(z_s\)
    \((\text{optical depth: }P\left(\text{SL}|z_s\right) \text{ or }\tau(z_s)\,)\).
    This probability can be recalculated for specified configurations of
    lens galaxies, leveraging \emph{multiprocessing} and \emph{njit}
    functionalities for enhanced efficiency.
  \item
    The package utilizes the Elliptical Power Law with external shear
    (EPL+Shear) model (Wempe et al. 2022) for galaxy parameters
    \((\theta_L)\) sampling, following Wierda et al. (2021). Rejection
    sampling is applied on the above samples on condition that whether
    the event is strongly lensed or not,
    \(P\left(\text{SL}|z_s,\theta_L\right)\).
  \end{itemize}
\item
  Generation of image properties:

  \begin{itemize}
  \tightlist
  \item
    Source position \((\beta)\) is sampled from the caustic in the
    source plane.
  \item
    Sampled lens properties and source position is fed in
    \emph{lenstronomy} (Birrer et al. 2021) to generate properties of
    the images. This is the slowest part of the entire simulation, which
    \(ler\) tackles through parallelization with multiprocessing.
  \item
    Image properties like magnification \((\mu_i)\) and time delay
    (\(\Delta t_i\)) modifies the original source signal strength,
    changing the signal-to-noise ratio SNR and our ability to detect.
  \end{itemize}
\item
  Calculation of Detectable Merger Rates Per Year:

  \begin{itemize}
  \tightlist
  \item
    The calculation of rates necessitates integration over simulated
    events that meet specific detection criteria. This process includes
    computing SNRs \((\rho)\) for each event or its lensed images,
    followed by an assessment against a predetermined threshold(s)
    \((\rho_{th})\).
  \item
    SNR calculations are optimized using
    \href{https://github.com/hemantaph/gwsnr}{\emph{gwsnr}}, leveraging
    interpolation and multiprocessing for accuracy and speed.
  \item
    Simulated events and rate results, along with input configurations,
    are systematically archived for easy access and future analysis.
    Additionally, all interpolators used in the process are preserved
    for future applications.
  \item
    Most cosmology-related calculations within the \emph{ler} package
    are performed using the \emph{astropy} library (Astropy
    Collaboration 2013). The default cosmological model is LambdaCDM
    (\(H_0=70,\, \Omega_m=0.3,\, \Omega_\Lambda=0.7\)); however, users
    have the flexibility to employ any cosmology available in
    \emph{astropy}. All internal calculations in \emph{ler} will then be
    based on the user-selected cosmological model.
  \end{itemize}
\end{itemize}

@@

\[
\begin{equation*}
\begin{split}
R_U = \int & dz_s \frac{dV_c}{dz_s}\frac{R_m(z_s)}{1+z_s}\left\{\Theta[\rho(z_s,\theta)-\rho_{th}] P(\theta) d\theta \right\}
\end{split}
\end{equation*}
\]

@@

\[
\begin{equation*}
\begin{split}
R_L = \int & dz_s \frac{dV_c}{dz_s}\tau(z_s)\frac{R_m(z_s)}{1+z_s} \,\mathcal{O}_{images}(z_s,\theta,\mu_i,\Delta t_i, \rho_{th}) \, \\ 
& \, P(\theta) P(\theta_L|\text{SL},z_s) P(\beta|\text{SL}) d\theta d\beta d\theta_L dz_s 
\end{split}
\end{equation*}
\]

@@

\phantomsection\label{refs}
\begin{CSLReferences}{1}{0}
\bibitem[\citeproctext]{ref-astropy}
Astropy Collaboration. 2013. {``{Astropy: A community Python package for
astronomy}.''} \emph{Astronomy and Astrophysics} 558 (October).
\url{https://doi.org/10.1051/0004-6361/201322068}.

\bibitem[\citeproctext]{ref-Birrer2021}
Birrer, Simon, Anowar J. Shajib, Daniel Gilman, Aymeric Galan, Jelle
Aalbers, Martin Millon, Robert Morgan, et al. 2021. {``Lenstronomy II: A
Gravitational Lensing Software Ecosystem.''} \emph{Journal of Open
Source Software} 6 (62): 3283.
\url{https://doi.org/10.21105/joss.03283}.

\bibitem[\citeproctext]{ref-gwcosmo}
gwcosmo Contributors. 2022. {``{gwcosmo: A Python package for
gravitational-wave cosmology}.''} \emph{GitHub Repository}. GitHub.
\url{https://github.com/gwcosmo/gwcosmo}.

\bibitem[\citeproctext]{ref-Janquart2023}
Janquart, J, M Wright, S Goyal, J C L Chan, A Ganguly, Á Garrón, D
Keitel, et al. 2023. {``{Follow-up analyses to the O3 LIGO--Virgo--KAGRA
lensing searches}.''} \emph{Monthly Notices of the Royal Astronomical
Society} 526 (3): 3832--60.
\url{https://doi.org/10.1093/mnras/stad2909}.

\bibitem[\citeproctext]{ref-numba}
Lam, Stan, Stéphane Pitrou, and Mark Seibert. 2022. {``Numba: A High
Performance Python Compiler.''} \emph{Numba Documentation}. Anaconda,
Inc. \url{https://numba.pydata.org/}.

\bibitem[\citeproctext]{ref-numpy}
NumPy Community. 2022. {``NumPy: A Fundamental Package for Scientific
Computing with Python.''} \emph{NumPy Website}. NumPy.
\url{https://numpy.org/}.

\bibitem[\citeproctext]{ref-scipy}
Virtanen, Pauli, Ralf Gommers, Travis E. Oliphant, Matt Haberland, Tyler
Reddy, David Cournapeau, Evgeni Burovski, et al. 2020. {``{SciPy 1.0:
Fundamental Algorithms for Scientific Computing in Python}.''}
\emph{Nature Methods}. SciPy. \url{https://www.scipy.org/}.
